---
title: "Problem Set 3"
author: "Walter Conrad"
date: "10/9/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

After determining the annual production and transportation costs for Supplier A and Supplier B, I have determined that Supplier B is the better choice. Although Supplier B's production cost is a bit higher than Supplier A's, shipping from Supplier B is much cheaper, and the overall total annual cost is much lower if Tool Depot chooses Supplier B.

I will explain the code that allowed me to perform this analysis.

First I ran the necessary libraries and imported the necessary data tables.

```{r}
library(tidyverse)
library(data.table)
library(gtable)
library(dplyr)
library(lubridate)
library(readxl)
```

```{r}
library(readxl)
Carrier_Rates <- read_excel("~/Desktop/Intro to Data Science/Homework/Homework 3/HW3.completed.summary.xlsx", 
    sheet = "Carrier Rates", col_types = c("text", 
        "text", "text", "numeric"))
```

```{r}
library(readxl)
Historical_Orders <- read_excel("~/Desktop/Intro to Data Science/Homework/Homework 3/HW3.completed.summary.xlsx", 
    sheet = "Historical Orders", col_types = c("date", 
        "text", "text", "text", "numeric", 
        "numeric", "numeric"))
```

```{r}
library(readxl)
Historical_Shipments <- read_excel("~/Desktop/Intro to Data Science/Homework/Homework 3/HW3.completed.summary.xlsx", 
    sheet = "Historical Shipments", col_types = c("date", 
        "text", "text", "numeric"))
```



Next I joined the data sets in order to perform analysis. First I left joined historical shipments with the carrier rates for carriers X and Y.

```{r}
shipments.by.rates <- left_join(Historical_Shipments,
         Carrier_Rates)

  shipments.by.rates
```

Next I joined the historical orders by the historical shipments using a left join.

```{r}
orders.by.shipments <- left_join(Historical_Orders,
                                 Historical_Shipments,
                                 by = c("Str Nbr",
                                        "Supplier",
                                        "Order Dt" = "Shipment Date"))

orders.by.shipments
```

Next I created the first of my "master" tables by joining both the previous two joined tables. This gave me enough information in one table to mutate the necessary columns for analysis.

```{r}
master <- left_join(orders.by.shipments,
                    shipments.by.rates,
                    by = c("Str Nbr",
                           "Supplier",
                           "Order Dt" = "Shipment Date"))

master
```

Next I created a "final.master" table. This table included the following mutated values: hammer_orders, real_hammer_orders,total_weight,ship_cost, and product_cost. Most of these values are self-explanatory. hammer_orders is a value derived from the number of wrench orders multiplied by a factor of 1.1 because hammer orders are expected to mirror wrench orders, and we expect a 10% increase in orders each year. real_hammer_orders simply applied the hammer_orders value to all origin-destination combinations (A1,A2,B1,B2) because in our data, store 1 is the only store that receives wrench shipments, but theoretically either store 1 or 2 from either Supplier A or B could receive hammers. total_weight is the total weight of a shipment for an origin-destination pair for a given week; this value is used to then calculate the ship_cost for that shipment in a given week. product_cost simply applies the product cost for Supplier A or B to the number of hammer orders.

```{r}
final.master <- master %>%
  group_by(week(`Order Dt`),
           `Str Nbr`) %>%
  mutate(hammer_orders = ifelse(`Product Id` == 1,
                                `Order Qty` * 1.1,
                                0)) %>%
  mutate(real_hammer_orders = max(hammer_orders)) %>%
  group_by(Carrier) %>%
  mutate(total_weight = 2 * real_hammer_orders + Weight.x) %>%
  mutate(ship_cost = ifelse(Carrier == "X", Cost, Cost * real_hammer_orders * 2 + Cost * Weight.x)) %>%
  mutate(product_cost = ifelse(Supplier == "A",
                             .80 * real_hammer_orders,
                             .82 * real_hammer_orders))

final.master
```


Using this final.master graph, I then filtered by Supplier and calculated minimum shipping costs for each week for both origin-destination combinations A1,A2 and B1,B2 respectively. Calculating product costs was as simple as summing the product cost for one supplier each week and dividing by a factor corresponding to the number of times the value was repeated each week in my table.

```{r}
#Annual Shipping Cost A
final.master %>%
  filter(Supplier == "A") %>%
  group_by(week(`Order Dt`),
           `Str Nbr`) %>%
  mutate(min_ship_cost = min(ship_cost)) %>%
  ungroup() %>%
  group_by(week(`Order Dt`)) %>%
  mutate(total_weekly_ship_cost = max(min_ship_cost)+min(min_ship_cost)) %>%
  ungroup() %>%
  summarize(annual_ship_cost = sum(total_weekly_ship_cost/8))
```
```{r}
#Annual product cost A
final.master %>%
  filter(Supplier == "A") %>%
  ungroup() %>%
  summarize(annual_product_cost = sum(product_cost/4))
```
```{r}
#Annual shipping cost B
final.master %>%
  filter(Supplier == "B") %>%
  group_by(week(`Order Dt`),
           `Str Nbr`) %>%
  mutate(min_ship_cost = min(ship_cost)) %>%
  ungroup() %>%
  group_by(week(`Order Dt`)) %>%
  mutate(total_weekly_ship_cost = max(min_ship_cost)+min(min_ship_cost)) %>%
  ungroup() %>%
  summarize(annual_ship_cost = sum(total_weekly_ship_cost/4))
```
```{r}
#Annual product cost B
final.master %>%
  filter(Supplier == "B") %>%
  ungroup() %>%
  summarize(annual_product_cost = sum(product_cost/2))
```

Using this information, I filled out the summary table and determined that Supplier B is the best choice for Tool Depot, mainly because the costs of shipping from Supplier B are substantially less than the cost to ship from Supplier A.


```{r}
library(readxl)
Summary <- read_excel("~/Desktop/Intro to Data Science/Homework/Homework 3/HW3.completed.summary.xlsx")

Summary
```



